The Evolution of Phenotypically Invariant Genetic Networks

### 2017-07-31 handnotes.tex
edited abstract

\begin{abstract}
 % \textbf{
We introduce an analytical theory to study the evolution of biological systems such as gene regulatory networks, applying insight and tools from control engineering, systems identification, and dynamical systems theory. We describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes – and the evolutionary paths connecting them. Under constant selection and environmental stasis, neutral evolution proceeds as a random walk through phenotypically identical network-space. Under neutral conditions, this model can describe the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift and speciation.
%}
\end{abstract}

>>>>

\begin{abstract}
Here we introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks. The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes. Here we give an exact expression for the set of all linear mechanisms with identical phenotypes. We employ a quantiative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit. We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
\end{abstract}
